# Independent factorial ANOVA {.section}

Two or more independent variables with two or more categories. One dependent variable.

```{r, echo=FALSE}
rm(list=ls())

data <- read.csv("anova_alcohol_speed_sekse.csv")

if(!"ggplot2" %in% installed.packages()) { install.packages("ggplot2") }
library("ggplot2")

# set.seed(1976)

if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Independent factorial ANOVA

The independent factorial ANOVA analyses the variance of multiple independent variables (Factors) with two or more categories.

Effects and interactions:

* 1 dependent/outcome variable
* 2 or more independent/predictor variables
    * 2 or more cat./levels
    
## Assumptions {.subsection}

* Continuous variable
* Random sample
* Normaly distributed 
    * Shapiro-Wilk test
* Equal variance within groups
    * Levene's test

## Formulas {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Model | $\text{SS}_{\text{model}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_{model}-1$ | $\frac{\text{SS}_{\text{model}}}{\text{df}_{\text{model}}}$ | $\frac{\text{MS}_{\text{model}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}A$ | $\text{SS}_{\text{A}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_A-1$ | $\frac{\text{SS}_{\text{A}}}{\text{df}_{\text{A}}}$ | $\frac{\text{MS}_{\text{A}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}B$ | $\text{SS}_{\text{B}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_B-1$ | $\frac{\text{SS}_{\text{B}}}{\text{df}_{\text{B}}}$ | $\frac{\text{MS}_{\text{B}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}AB$ | $\text{SS}_{A \times B} = \text{SS}_{\text{model}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}}$ | $df_A \times df_B$ | $\frac{\text{SS}_{\text{AB}}}{\text{df}_{\text{AB}}}$ | $\frac{\text{MS}_{\text{AB}}}{\text{MS}_{\text{error}}}$ |
| Error  | $\text{SS}_{\text{error}} = \sum{s_k^2(n_k-1)}$ | $N-k_{model}$ | $\frac{\text{SS}_{\text{error}}}{\text{df}_{\text{error}}}$ | |
| Total | $\text{SS}_{\text{total}} = \text{SS}_{\text{model}} + \text{SS}_{\text{error}}$ | $N-1$ | $\frac{\text{SS}_{\text{total}}}{\text{df}_{\text{total}}}$ | |

## Example {.smaller}

In this example we will look at the amount of accidents in a car driving simulator while subjects where given varying doses of speed and alcohol.

* Dependent variable
    * Accidents
* Independent variables
    * Speed
        * None
        * Small
        * Large
    * Alcohol
        * None
        * Small
        * Large
        
## {.smaller}

person | alcohol | speed | accidents
-------|---------|-------|----------
1      | 1       | 1     | 0
2      | 1       | 2     | 2
3      | 1       | 3     | 4
4      | 2       | 1     | 6
5      | 2       | 2     | 8
6      | 2       | 3     | 10
7      | 3       | 1     | 12
8      | 3       | 2     | 14
9      | 3       | 3     | 16
      
## Data {.smaller}

```{r, echo=FALSE}
datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## SS<sub>model</sub> {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Model | $\text{SS}_{\text{model}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_{model}-1$ | $\frac{\text{SS}_{\text{model}}}{\text{df}_{\text{model}}}$ | $\frac{\text{MS}_{\text{model}}}{\text{MS}_{\text{error}}}$ |

<br>

```{r, echo=FALSE}
accidents <- data$accidents
n         <- length(accidents)

# Descriptives for model
means.model <- aggregate(accidents ~ speed + alcohol + speed * alcohol, data, mean  )
n.k.model   <- aggregate(accidents ~ speed + alcohol + speed * alcohol, data, length)

cbind(means.model, n = n.k.model$accidents)
```

--------

```{r, echo=FALSE}
# Calculate expected accidents
fit <- lm(accidents ~ speed + alcohol + speed * alcohol, data)
exp.accidents <- fit$fitted.values

datatable(data.frame(data$speed, 
                     data$alcohol, 
                     accidents = round(accidents,2), 
                     grand.mean = round(mean(accidents),2), 
                     group.mean = round(exp.accidents,2)), options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

--------

```{r, echo=TRUE}
SS.model <- sum((exp.accidents - mean(data$accidents))^2); SS.model
```

```{r}
m.k1 = mean(subset(data, speed == "none" & alcohol == "none", select = "accidents")$accidents)
m.k2 = mean(subset(data, speed == "none" & alcohol == "some", select = "accidents")$accidents)
m.k3 = mean(subset(data, speed == "none" & alcohol == "much", select = "accidents")$accidents)
m.k4 = mean(subset(data, speed == "some" & alcohol == "none", select = "accidents")$accidents)
m.k5 = mean(subset(data, speed == "some" & alcohol == "some", select = "accidents")$accidents)
m.k6 = mean(subset(data, speed == "some" & alcohol == "much", select = "accidents")$accidents)
m.k7 = mean(subset(data, speed == "much" & alcohol == "none", select = "accidents")$accidents)
m.k8 = mean(subset(data, speed == "much" & alcohol == "some", select = "accidents")$accidents)
m.k9 = mean(subset(data, speed == "much" & alcohol == "much", select = "accidents")$accidents)

n.k1 = n.k2 = n.k3 = n.k4 = n.k5 = n.k6 = n.k7 = n.k8 = n.k9 = 20
```

---------

```{r}
ss.m.k1 = sum(n.k1 * (m.k1 - mean(accidents))^2 )
ss.m.k2 = sum(n.k2 * (m.k2 - mean(accidents))^2 )
ss.m.k3 = sum(n.k3 * (m.k3 - mean(accidents))^2 )
ss.m.k4 = sum(n.k4 * (m.k4 - mean(accidents))^2 )
ss.m.k5 = sum(n.k5 * (m.k5 - mean(accidents))^2 )
ss.m.k6 = sum(n.k6 * (m.k6 - mean(accidents))^2 )
ss.m.k7 = sum(n.k7 * (m.k7 - mean(accidents))^2 )
ss.m.k8 = sum(n.k8 * (m.k8 - mean(accidents))^2 )
ss.m.k9 = sum(n.k9 * (m.k9 - mean(accidents))^2 )

ss.model = sum(ss.m.k1,ss.m.k2,ss.m.k3,ss.m.k4,ss.m.k5,ss.m.k6,ss.m.k7,ss.m.k8,ss.m.k9)
ss.model
```

## SS<sub>model</sub> visual {.smaller}

```{r}
# Plot all data points
plot(accidents,
     xlab = 'participants')

# With the mean
lines(1:n,rep(mean(accidents),n),col='black',lwd=2)

# The black lines are the total variance, the deviation from the mean.
segments(1:n, mean(accidents), 1:n, accidents)

# The model predicts the accidents scores
points(1:n,exp.accidents, col='red')

p <- recordPlot()

# The black lines are the total variance, the deviation from the mean.
segments(1:n, exp.accidents, 1:n, mean(accidents), col = "red")

# Add legend to plot
legend("topleft",
       pch    = c("o"),
       col    = c("red"),
       legend = c("Full model") )
```

## SS<sub>error</sub> {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Error  | $\text{SS}_{\text{error}} = \sum{s_k^2(n_k-1)}$ | $N-k$ | $\frac{\text{SS}_{\text{error}}}{\text{df}_{\text{error}}}$ | |

<br>

```{r, echo=FALSE, eval=FALSE}
SS.error <- sum((accidents - exp.accidents)^2)
SS.error
```

```{r}
v.k1 = var(subset(data, speed == "none"   & alcohol == "none",   select = "accidents")$accidents)
v.k2 = var(subset(data, speed == "none"   & alcohol == "some", select = "accidents")$accidents)
v.k3 = var(subset(data, speed == "none"   & alcohol == "much",   select = "accidents")$accidents)
v.k4 = var(subset(data, speed == "some" & alcohol == "none",   select = "accidents")$accidents)
v.k5 = var(subset(data, speed == "some" & alcohol == "some", select = "accidents")$accidents)
v.k6 = var(subset(data, speed == "some" & alcohol == "much",   select = "accidents")$accidents)
v.k7 = var(subset(data, speed == "much"   & alcohol == "none",   select = "accidents")$accidents)
v.k8 = var(subset(data, speed == "much"   & alcohol == "some", select = "accidents")$accidents)
v.k9 = var(subset(data, speed == "much"   & alcohol == "much",   select = "accidents")$accidents)
```

---------

```{r}
ss.e.k1 = v.k1 * (n.k1 - 1)
ss.e.k2 = v.k2 * (n.k2 - 1)
ss.e.k3 = v.k3 * (n.k3 - 1)
ss.e.k4 = v.k4 * (n.k4 - 1)
ss.e.k5 = v.k5 * (n.k5 - 1)
ss.e.k6 = v.k6 * (n.k6 - 1)
ss.e.k7 = v.k7 * (n.k7 - 1)
ss.e.k8 = v.k8 * (n.k8 - 1)
ss.e.k9 = v.k9 * (n.k9 - 1)

ss.error = sum(ss.e.k1,ss.e.k2,ss.e.k3,ss.e.k4,ss.e.k5,ss.e.k6,ss.e.k7,ss.e.k8,ss.e.k9)
ss.error
```

## SS<sub>error</sub> visual

```{r}
p

# Now we can add the error of this model
segments(1:n, accidents, 1:n, exp.accidents, col='purple')

# Add legend to plot
legend("topleft",
       pch    = c("o", "-"),
       col    = c("red", "purple"),
       legend = c("Full model", "error") )
```

## SS<sub>A</sub> Speed {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}A$ | $\text{SS}_{\text{A}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_A-1$ | $\frac{\text{SS}_{\text{A}}}{\text{df}_{\text{A}}}$ | $\frac{\text{MS}_{\text{A}}}{\text{MS}_{\text{error}}}$ |

<br>

```{r, echo=FALSE}
# Descriptives for speed
means.speed <- aggregate(accidents ~ speed, data, mean  )
n.k.speed   <- aggregate(accidents ~ speed, data, length)
data$mean.total <- mean(data$accidents)
names(means.speed)[2] <- "means.speed"
#means.speed

data <- merge(data, means.speed,  by="speed",  all.x=TRUE)
data <- data[order(data$subjects),]
```

```{r, echo = FALSE}
SS.speed <- sum(with(data, (means.speed - mean.total)^2))
#SS.speed
```

```{r}
m.s1 = mean(subset(data, speed == "none", select = "accidents")$accidents)
m.s2 = mean(subset(data, speed == "some", select = "accidents")$accidents)
m.s3 = mean(subset(data, speed == "much", select = "accidents")$accidents)

n.s1 = n.s2 = n.s3 = 60
```

-------

```{r}
ss.s1 = sum(n.s1 * (m.s1 - mean(accidents))^2 )
ss.s2 = sum(n.s2 * (m.s2 - mean(accidents))^2 )
ss.s3 = sum(n.s3 * (m.s3 - mean(accidents))^2 )

ss.speed = sum(ss.s1,ss.s2,ss.s3)
ss.speed
```

## SS<sub>A</sub> Speed Visual

```{r}
p
# Deviation of speed
points(1:n, data$means.speed,  col='green')
segments(1:n, mean(accidents), 1:n, data$means.speed, col='green')

# Add legend to plot
legend("topleft",
       pch    = c("o", "o"),
       col    = c("green", "red"),
       legend = c("speed", "Full model") )
```

## SS<sub>B</sub> Alcohol {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}B$ | $\text{SS}_{\text{B}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_B-1$ | $\frac{\text{SS}_{\text{B}}}{\text{df}_{\text{B}}}$ | $\frac{\text{MS}_{\text{B}}}{\text{MS}_{\text{error}}}$ |

<br>

```{r, echo=FALSE}
# Descriptives for alcohol
means.alcohol <- aggregate(accidents ~ alcohol, data, mean)
n.k.alcohol   <- aggregate(accidents ~ alcohol, data, length)
names(means.alcohol)[2] <- "means.alcohol"
# means.alcohol

data <- merge(data,means.alcohol,by="alcohol",all.x=TRUE)
data <- data[order(data$subjects),]
```

```{r, echo=FALSE}
SS.alcohol <- sum(with(data, (means.alcohol - mean.total)^2))
# SS.alcohol
```

```{r}
m.a1 = mean(subset(data, alcohol == "none", select = "accidents")$accidents)
m.a2 = mean(subset(data, alcohol == "some", select = "accidents")$accidents)
m.a3 = mean(subset(data, alcohol == "much", select = "accidents")$accidents)

n.a1 = n.a2 = n.a3 = 60
```

------

```{r}
ss.a1 = sum(n.a1 * (m.a1 - mean(accidents))^2 )
ss.a2 = sum(n.a2 * (m.a2 - mean(accidents))^2 )
ss.a3 = sum(n.a3 * (m.a3 - mean(accidents))^2 )

ss.alcohol = sum(ss.a1,ss.a2,ss.a3)
ss.alcohol
```

## SS<sub>B</sub> Alcohol Visual

```{r}
p

# Deviation of alcohol
points(1:n, data$means.alcohol,col='orange',type="p")
segments(1:n, mean(accidents), 1:n, data$means.alcohol, col='orange')

# Add legend to plot
legend("topleft",
       pch    = c("o", "o"),
       col    = c("orange", "red"),
       legend = c("alcohol", "Full model") )
```

## SS<sub>AB</sub> Alcohol x Speed {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}AB$ | $\text{SS}_{A \times B} = \text{SS}_{\text{model}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}}$ | $df_A \times df_B$ | $\frac{\text{SS}_{\text{AB}}}{\text{df}_{\text{AB}}}$ | $\frac{\text{MS}_{\text{AB}}}{\text{MS}_{\text{error}}}$ |

<br>

```{r}
# Sums of squares for the interaction between speed and alcohol
ss.speed.alcohol <- ss.model - ss.speed - ss.alcohol
ss.speed.alcohol
```

## Mean Squares

Mean squares for:

* Speed
* Alcohol
* Speed $\times$ Alcohol

$$\begin{aligned}
F_{Speed} &= \frac{{MS}_{Speed}}{{MS}_{error}} \\
F_{Alcohol} &= \frac{{MS}_{Alcohol}}{{MS}_{error}} \\
F_{Alcohol \times Speed} &= \frac{{MS}_{Alcohol \times Speed}}{{MS}_{error}} \\
\end{aligned}$$

## Interaction

$$F_{Alcohol \times Speed}$$

```{r}
N          = length(accidents)
k.speed    = 3
k.alcohol  = 3
k.model    = 9
df.speed   = k.speed   - 1
df.alcohol = k.alcohol - 1
df.speed.alcohol = df.speed * df.alcohol

ms.speed.alcohol = ss.speed.alcohol / df.speed.alcohol

df.error = N - k.model
ms.error = ss.error / df.error
```

-----

```{r}
F.speed.alcohol = ms.speed.alcohol / ms.error
F.speed.alcohol

library(visualize)
visualize.f(F.speed.alcohol, df.speed.alcohol, df.error, section = "upper")
```

------

```{r}
F.values = seq(0, 10, .01)

plot(F.values, df(F.values, df.speed.alcohol, df.error), type = "l", ylab="density", main = "H0 and HA F-distribution")

critical.value = qf(.95, df.speed.alcohol, df.error)

critical.range = seq(critical.value, 10, .01)

polygon(c(critical.range,rev(critical.range)), 
        c(critical.range*0, rev(df(critical.range, df.speed.alcohol, df.error, ncp = 5))), col = "green")

lines(F.values, df(F.values, df.speed.alcohol, df.error, ncp = 5))

polygon(c(critical.range,rev(critical.range)), 
        c(critical.range*0, rev(df(critical.range, df.speed.alcohol, df.error))), col = "red")

p.range = seq(F.speed.alcohol, 10, .01)

polygon(c(p.range,rev(p.range)), 
        c(p.range*0, rev(df(p.range, df.speed.alcohol, df.error))), col = rgb(0,0,1,.5), density = 40)
```

## Contrast

Planned comparisons

* Exploring differences of theoretical interest
* Higher precision
* Higher power

## Post-Hoc

Unplanned comparisons

* Exploring all possible differences
* Adjust T value for inflated type 1 error

## Effect size

General effect size measures

* Amount of explained variance $R^2$ also called eta squared $\eta^2$.
* Omega squared $\omega^2$

Effect sizes of contrasts or post-hoc comparisons

* Cohen's $r$ gives the effect size for a specific comparison
    * $r_{Contrast} = \sqrt{\frac{t^2}{t^2+{df}}}$
