# Independent factorial ANOVA {.section}

Two or more independent variables with two or more categories. One dependent variable.

```{r, echo=FALSE}
rm(list=ls())

data <- read.csv("anova_alcohol_speed_sekse.csv")

if(!"ggplot2" %in% installed.packages()) { install.packages("ggplot2") }
library("ggplot2")

set.seed(1976)

if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Independent factorial ANOVA

The independent factorial ANOVA analyses the variance of multiple independent variables with two or more categories (Factors). 

## Formulas {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Model | $\text{SS}_{\text{model}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_{model}-1$ | $\frac{\text{SS}_{\text{model}}}{\text{df}_{\text{model}}}$ | $\frac{\text{MS}_{\text{model}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}A$ | $\text{SS}_{\text{A}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_A-1$ | $\frac{\text{SS}_{\text{A}}}{\text{df}_{\text{A}}}$ | $\frac{\text{MS}_{\text{A}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}B$ | $\text{SS}_{\text{B}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_B-1$ | $\frac{\text{SS}_{\text{B}}}{\text{df}_{\text{B}}}$ | $\frac{\text{MS}_{\text{B}}}{\text{MS}_{\text{error}}}$ |
| $\hspace{2ex}AB$ | $\text{SS}_{A \times B} = \text{SS}_{\text{model}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}}$ | $df_A \times df_B$ | $\frac{\text{SS}_{\text{AB}}}{\text{df}_{\text{AB}}}$ | $\frac{\text{MS}_{\text{AB}}}{\text{MS}_{\text{error}}}$ |
| Error  | $\text{SS}_{\text{error}} = \sum{s_k^2(n_k-1)}$ | $N-k$ | $\frac{\text{SS}_{\text{error}}}{\text{df}_{\text{error}}}$ | |
| Total | $\text{SS}_{\text{total}} = \text{SS}_{\text{model}} + \text{SS}_{\text{error}}$ | $N-1$ | $\frac{\text{SS}_{\text{total}}}{\text{df}_{\text{total}}}$ | |

## Example {.smaller}

In this example we will look at the amount of accidents in a car driving simulator while subjects where given varying doses of speed and alcohol.

* Dependent variable
    * Accidents
* Independent variables
    * Speed
        * None
        * Small
        * Large
    * Alcohol
        * None
        * Small
        * Large
      
## Data {.smaller}

```{r, echo=FALSE}
datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## SS<sub>model</sub> {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Model | $\text{SS}_{\text{model}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_{model}-1$ | $\frac{\text{SS}_{\text{model}}}{\text{df}_{\text{model}}}$ | $\frac{\text{MS}_{\text{model}}}{\text{MS}_{\text{error}}}$ |

```{r, echo=FALSE}
brokken <- data$brokken
n       = length(brokken)

# Descriptives for model
means.model <- aggregate(brokken ~ speed + alcohol + speed * alcohol, data, mean  )
n.k.model   <- aggregate(brokken ~ speed + alcohol + speed * alcohol, data, length)

cbind(means.model, n.k.model$brokken)
```

--------

```{r, echo=FALSE}
# Calculate expected brokken
fit <- lm(brokken ~ speed + alcohol + speed * alcohol, data)
exp.brokken <- fit$fitted.values

datatable(data.frame(data$speed, 
                     data$alcohol, 
                     brokken = round(brokken,2), 
                     mean = round(mean(brokken),2), 
                     exp.brokken = round(exp.brokken,2)), options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

--------

```{r}
SS.model <- sum((exp.brokken - mean(data$brokken))^2)
SS.model
```

## SS<sub>model</sub> visual {.smaller}

```{r}
# Plot all data points
plot(brokken,
     xlab = 'participants')

# With the mean
lines(1:n,rep(mean(brokken),n),col='black',lwd=2)

# The black lines are the total variance, the deviation from the mean.
segments(1:n, mean(brokken), 1:n, brokken)

# The model predicts the brokken scores
points(1:n,exp.brokken, col='red')

p <- recordPlot()

# The black lines are the total variance, the deviation from the mean.
segments(1:n, exp.brokken, 1:n, mean(brokken), col = "red")

# Add legend to plot
legend("topleft",
       pch    = c("o"),
       col    = c("red"),
       legend = c("Full model") )
```

## SS<sub>error</sub> {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| Error  | $\text{SS}_{\text{error}} = \sum{s_k^2(n_k-1)}$ | $N-k$ | $\frac{\text{SS}_{\text{error}}}{\text{df}_{\text{error}}}$ | |

```{r}
SS.error <- sum((brokken - exp.brokken)^2)
SS.error
```

## SS<sub>error</sub> visual

```{r}
p

# Now we can add the error of this model
segments(1:n, brokken, 1:n, exp.brokken, col='purple')

# Add legend to plot
legend("topleft",
       pch    = c("o", "-"),
       col    = c("red", "purple"),
       legend = c("Full model", "error") )
```

## SS<sub>A</sub> Speed {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}A$ | $\text{SS}_{\text{A}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_A-1$ | $\frac{\text{SS}_{\text{A}}}{\text{df}_{\text{A}}}$ | $\frac{\text{MS}_{\text{A}}}{\text{MS}_{\text{error}}}$ |

```{r, echo=FALSE}
# Descriptives for speed
means.speed <- aggregate(brokken ~ speed, data, mean  )
n.k.speed   <- aggregate(brokken ~ speed, data, length)
data$mean.total <- mean(data$brokken)
names(means.speed)[2] <- "means.speed"
means.speed

data <- merge(data, means.speed,  by="speed",  all.x=TRUE)
data <- data[order(data$pp),]
```

```{r}
SS.speed <- sum(with(data, (means.speed   - mean.total)^2))
SS.speed
```

## SS<sub>A</sub> Speed Visual

```{r}
p
# Deviation of speed
points(1:n, data$means.speed,  col='green')
segments(1:n, mean(brokken), 1:n, data$means.speed, col='green')

# Add legend to plot
legend("topleft",
       pch    = c("o", "o"),
       col    = c("green", "red"),
       legend = c("speed", "Full model") )
```

## SS<sub>B</sub> Alcohol {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}B$ | $\text{SS}_{\text{B}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k_B-1$ | $\frac{\text{SS}_{\text{B}}}{\text{df}_{\text{B}}}$ | $\frac{\text{MS}_{\text{B}}}{\text{MS}_{\text{error}}}$ |

```{r, echo=FALSE}
# Descriptives for alcohol
means.alcohol <- aggregate(brokken ~ alcohol, data, mean)
n.k.alcohol   <- aggregate(brokken ~ alcohol, data, length)
names(means.alcohol)[2] <- "means.alcohol"
means.alcohol

data <- merge(data,means.alcohol,by="alcohol",all.x=TRUE)
data <- data[order(data$pp),]
```

```{r}
SS.alcohol <- sum(with(data, (means.alcohol - mean.total)^2))
SS.alcohol
```

## SS<sub>B</sub> Alcohol Visual

```{r}
p

# Deviation of alcohol
points(1:n, data$means.alcohol,col='orange',type="p")
segments(1:n, mean(brokken), 1:n, data$means.alcohol, col='orange')

# Add legend to plot
legend("topleft",
       pch    = c("o", "o"),
       col    = c("orange", "red"),
       legend = c("alcohol", "Full model") )
```

## SS<sub>AB</sub> Alcohol x Speed {.subsection}

| Variance | Sum of squares | df | Mean squares | F-ratio |
|----------|----------------|----|--------------|---------|
| $\hspace{2ex}AB$ | $\text{SS}_{A \times B} = \text{SS}_{\text{model}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}}$ | $df_A \times df_B$ | $\frac{\text{SS}_{\text{AB}}}{\text{df}_{\text{AB}}}$ | $\frac{\text{MS}_{\text{AB}}}{\text{MS}_{\text{error}}}$ |

```{r}
# Sums of squares for the interaction between speed and alcohol

SS.speed.alcohol <- SS.model - SS.speed - SS.alcohol
SS.speed.alcohol
```