# Nonparametric tests

## Parametric vs Nonparametric {#mytable}

<style>
#mytable th {
background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #FFA500), color-stop(80%, #FFA500)) no-repeat !important;
background-image: -webkit-linear-gradient(top, rgb(255,165,0) 40%, rgb(255,165,0) 80%) !important;
background: -webkit-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background-image: -webkit-linear-gradient(top, rgb(255,165,0) 40%, rgb(255,165,0) 80%) !important;
background: -moz-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background: -o-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background: linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
}
</style>

Attribute    | Parametric            | Nonparametric
-------------|-----------------------|--------------------------
distribution | normaly distributed   | any distribution
sampling     | random sample         | random sample
sensitivity to outliers | yes        | no
works with   | large data sets       | small and large data sets
speed        | fast                  | slow

## Ranking

```{r, fig.height=3, fig.align='center'}
x = c(1, 4, 6, 7, 8, 9); y = c(1, 4, 6, 7, 8, 39)

layout(matrix(1:2, 1, 2))
boxplot(x, horizontal=T, col='orange'); boxplot(y, horizontal=T, col='orange')
```

## Ranking

```{r, echo=FALSE, fig.height=3, fig.align='center'}
layout(matrix(1:2, 1, 2))
boxplot(rank(x), horizontal=T, col='orange'); boxplot(rank(y), horizontal=T, col='orange')
```

```{r, fig.height=3}
rbind(rx = rank(x), ry = rank(y))
```

## Ties

```{r}
x = c(1, 4, 6, 7, 8, 8)
rank(x)
```

$$\frac{5 + 6}{2} = 5.5$$

```{r}
rbind(x, Rx = rank(x))
```


## Procedure

1. Assumption: independent random samples.
2. Hypothesis: 
<br>$H_0$ : equal population distributions (implies equal mean ranking) 
<br>$H_A$ : unequal mean ranking (two sided) 
<br>$H_A$ : higher mean ranking for one group.
3. Test statistic is difference between mean or sum of ranking.
4. Standardise test statistic
5. Calculate *P*-value one or two sided.
5. Conclude to reject $H_0$ if $p < \alpha$.