# Paired-samples t-test {.section}

Paired 2 samples

## Paired-samples t-test

In the Paired samples t-test the deviation ($D$) for each pair is calculated and the mean of these deviations ($\bar{D}$) is tested against the null hypothesis where $\mu = 0$.

$$t_{n-1} = \frac{\bar{D} - \mu}{ {SE}_D }$$
Where $n$ (the number of cases) minus $1$, are the degrees of freedom $df = n - 1$ and $SE_D$ is the standard error of $D$, defined as $s_D/\sqrt{n}$.

## Hypothesis

$$\LARGE{
  \begin{aligned}
  H_0 &: \bar{D} = \mu_D \\
  H_A &: \bar{D} \neq \mu_D \\
  H_A &: \bar{D} > \mu_D \\
  H_A &: \bar{D} < \mu_D \\
  \end{aligned}}$$

## Data structure

index | k1 | k2
------|----|--
1     | d  | d
2     | d  | d
3     | d  | d
4     | d  | d

Where $k$ is the level of the categorical predictor variabla and $d$ is the value of the outcome/dependent variable.

## Data example

We are going to use the IQ estimates we collected last week. You had to gues the IQ of the one sitting next to you and your own IQ.

Let's take a look at the data.

## IQ estimates

```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(7845)

# Remove all objects from workspace
rm(list=ls())

# Install if nesserary and load RCurl package
if(!'RCurl' %in% installed.packages()) { install.packages('RCurl') }
library(RCurl)

# Load csv
myCsv <- getURL("https://docs.google.com/spreadsheets/d/13oLFjy5e923weH8bzBipZSvWYS_dpw7B77g5yW_myLg/export?format=csv&id=13oLFjy5e923weH8bzBipZSvWYS_dpw7B77g5yW_myLg")
# Read csv and assing to data object
data <- read.csv(textConnection(myCsv))

data = data[grep("2017", as.character(data$Timestamp)),c("IQ.van.je.buur", "Eigen.IQ")]

IQ.next.to.you = data$IQ.van.je.buur
IQ.you           = data$Eigen.IQ

if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

```{r, echo=FALSE}
datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Calculate $D$ {.smaller .subsection}

```{r}
D = IQ.next.to.you - IQ.you
```


```{r, echo=FALSE}
data$D = D

datatable(data, options = list(searching = FALSE, scrollY = 315, paging = F, info = F))
```

## Calculate $\bar{D}$

```{r}
D      = na.omit(D) # get rid of all missing values
D.mean = mean(D)
D.mean
```

And we also need n.

```{r}
n = length(D)
n
```

## Calculate t-value {.subsection}

$$t_{n-1} = \frac{\bar{D} - \mu}{ {SE}_D }$$

```{r}
mu = 0                # Define mu

D.sd = sd(D)          # Calculate standard deviation
D.se = D.sd / sqrt(n) # Calculate standard error

df   = n - 1          # Calculate degrees of freedom

# Calculate t
t = ( D.mean - mu ) / D.se
t
```

## Test for significance {.subsection}

Two tailed

```{r}
if(!"visualize" %in% installed.packages()) { install.packages("visualize") }
library("visualize")

visualize.t(c(-t,t), df, section="tails")
```

## Effect-size {.subsection}

$$r = \sqrt{\frac{t^2}{t^2 + \text{df}}}$$

```{r}
r = sqrt(t^2/(t^2 + df))
r
```

## Confidence interval {.smaller .subsection}

To display correct conficance intervals in SPSS we need to correct the original scores for whithin subject variation.

```{}
** SPSS SYNTAX

COMPUTE personal_mean = MEAN(IQ.next.to.you, IQ.you).
EXECUTE.

AGGREGATE
  /OUTFILE=* MODE=ADDVARIABLES
  /BREAK=
  /total_mean = MEAN(personal_mean).

COMPUTE adjustment = total_mean - personal_mean.
EXECUTE.

COMPUTE IQ.next.to.you.adj = IQ.next.to.you + adjustment.
COMPUTE IQ.you             = IQ.you         + adjustment.
EXECUTE.
```