# ANOVA factorial repeated

```{r, echo=FALSE}
rm(list=ls())
if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Factorial repeated measures ANOVA {.section}

The factorial repeated measures ANOVA analyses the variance of the model while reducing the error by the within person variance.

* 1 dependent/outcome variable
* 2 or more independent/predictor variable
    * 2 or more levels
* All with same subjects

## Assumptions

Same as one-way repeated measures ANOVA

## Example {.smaller}

In this example we will again look at the amount of accidents in a car driving simulator while subjects where given varying doses of speed and alcohol. But this time we lat participants partake in all conditions. Every week subjects returned for a different experimental condition.

* Dependent variable
    * Accidents
* Independent variables
    * Speed
        * None
        * Small
        * Large
    * Alcohol
        * None
        * Small
        * Large
        
## {.smaller}

person | 1_1 | 1_2 | 1_3 | 2_1 | 2_2 | 2_3 | 3_1 | 3_2 | 3_3
-------|-----|-----|-----|-----|-----|-----|-----|-----|----
1      | 1   |     |     |     |     |     |     |     |
2      |     | 2   |     |     |     |     |     |     |
3      |     |     | 3   |     |     |     |     |     |
4      |     |     |     | 4   |     |     |     |     |
5      |     |     |     |     | 5   |     |     |     |
6      |     |     |     |     |     | 6   |     |     |
7      |     |     |     |     |     |     | 7   |     |
8      |     |     |     |     |     |     |     | 8   |
9      |     |     |     |     |     |     |     |     | 9

## Data {.smaller}

```{r, echo=FALSE}
data <- read.csv(file = "ANOVA REP alcohol_speed.csv", header = T)

data[,-1] <- round(data[,-1],2)

datatable(data, rownames = FALSE, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```


