# Mixed design ANOVA {.section}

```{r, echo=FALSE}
rm(list=ls())
if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Mixed design 

The mixed ANOVA analyses the variance of the model while reducing the error by the within person variance.

* 1 dependent/outcome variable
* 2 or more independent/predictor variable with different subjects
    * 2 or more levels
* 2 or more independent/predictor variable with same subjects
    * 2 or more levels

## Assumptions

Same as repeated measures ANOVA and same as factorial ANOVA.

## Example {.smaller}

<small>

* Dependent variable
    * Accidents
* Independent variables
    * Speed (same subjects)
        * None
        * Small
        * Large
    * Alcohol (same subjects)
        * None
        * Small
        * Large
    * Gender
        * Males
        * Females

</small>
        
## {.smaller}

<small>

person | gender  | 1_1 | 1_2 | 1_3 | 2_1 | 2_2 | 2_3 | 3_1 | 3_2 | 3_3
-------|---------|-----|-----|-----|-----|-----|-----|-----|-----|----
1      | males   | 1   |     |     |     |     |     |     |     |
2      | males   |     | 2   |     |     |     |     |     |     |
3      | males   |     |     | 3   |     |     |     |     |     |
4      | males   |     |     |     | 4   |     |     |     |     |
5      | males   |     |     |     |     | 5   |     |     |     |
6      | males   |     |     |     |     |     | 6   |     |     |
7      | males   |     |     |     |     |     |     | 7   |     |
8      | males   |     |     |     |     |     |     |     | 8   |
9      | males   |     |     |     |     |     |     |     |     | 9
10     | females | 1   |     |     |     |     |     |     |     |
12     | females |     | 2   |     |     |     |     |     |     |
13     | females |     |     | 3   |     |     |     |     |     |
14     | females |     |     |     | 4   |     |     |     |     |
15     | females |     |     |     |     | 5   |     |     |     |
16     | females |     |     |     |     |     | 6   |     |     |
17     | females |     |     |     |     |     |     | 7   |     |
18     | females |     |     |     |     |     |     |     | 8   |
20     | females |     |     |     |     |     |     |     |     | 9

</small>

## Data {.smaller}

```{r, echo=FALSE}
data <- read.csv(file = "ANOVA_mixed.csv", header = T)

data[,-1] <- round(data[,-1],2)

datatable(data, rownames = FALSE, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Reshape data

```{r}
data.long <- reshape(data,
                     idvar     = "pp",
                     varying   = list(3:11),
                     v.names   = "Accidents",
                     direction = "long")

data.long[which(data.long$time == 1), "time"] = "1_1"
data.long[which(data.long$time == 2), "time"] = "1_2"
data.long[which(data.long$time == 3), "time"] = "1_3"
data.long[which(data.long$time == 4), "time"] = "2_1"
data.long[which(data.long$time == 5), "time"] = "2_2"
data.long[which(data.long$time == 6), "time"] = "2_3"
data.long[which(data.long$time == 7), "time"] = "3_1"
data.long[which(data.long$time == 8), "time"] = "3_2"
data.long[which(data.long$time == 9), "time"] = "3_3"

data.long$speed = NA
data.long$speed[grep("^1", data.long$time)] = "none"
data.long$speed[grep("^2", data.long$time)] = "small"
data.long$speed[grep("^3", data.long$time)] = "large"

data.long$alcohol = NA
data.long$alcohol[grep("_1$", data.long$time)] = "none"
data.long$alcohol[grep("_2$", data.long$time)] = "small"
data.long$alcohol[grep("_3$", data.long$time)] = "large"

data.long$pp      <- as.factor(data.long$pp)
data.long$sekse   <- as.factor(data.long$sekse)
data.long$speed   <- as.factor(data.long$speed)
data.long$alcohol <- as.factor(data.long$alcohol)
```

## Data long

```{r}
data.long
```

## ezANOVA

```{r}
if(!"ez" %in% installed.packages()) { install.packages("ez") }
library("ez")

Accidents.aov <- ezANOVA(data       = data.long, 
                         dv         = Accidents, 
                         wid        = pp, 
                         within     = c(speed,alcohol), 
                         between    = sekse, 
                         type       = 3, 
                         return_aov = TRUE)
```

## Results

```{r}
Accidents.aov
```
## Mauchly's Test for Sphericity

```{r}
Accidents.aov$`Mauchly's Test for Sphericity`
```

## Visualize results

```{r}
ezPlot(data    = data.long,
       x       = .(speed),
       col     = .(sekse),
       split   = .(alcohol),
       dv      = .(Accidents), 
       wid     = .(pp), 
       within  = .(speed,alcohol), 
       between = .(sekse) 
       )


```

