# Mixed design ANOVA {.section}

```{r, echo=FALSE}
rm(list=ls())
if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Mixed design 

The mixed ANOVA analyses the variance of the model while reducing the error by the within person variance.

* 1 dependent/outcome variable
* 2 or more independent/predictor variable with different subjects
    * 2 or more levels
* 1 or more independent/predictor variable with same subjects
    * 2 or more levels

## Assumptions

Same as repeated measures ANOVA and same as factorial ANOVA.

## Example {.smaller}

<small>

* Dependent variable
    * Accidents
* Independent variables
    * Speed (same subjects)
        * None
        * Small
        * Large
    * Alcohol (same subjects)
        * None
        * Small
        * Large
    * Gender
        * Males
        * Females

</small>
        
## {.smaller}

<small>

person | gender  | 1_1 | 1_2 | 1_3 | 2_1 | 2_2 | 2_3 | 3_1 | 3_2 | 3_3
-------|---------|-----|-----|-----|-----|-----|-----|-----|-----|----
1      | males   | 1   |     |     |     |     |     |     |     |
2      | males   |     | 2   |     |     |     |     |     |     |
3      | males   |     |     | 3   |     |     |     |     |     |
4      | males   |     |     |     | 4   |     |     |     |     |
5      | males   |     |     |     |     | 5   |     |     |     |
6      | males   |     |     |     |     |     | 6   |     |     |
7      | males   |     |     |     |     |     |     | 7   |     |
8      | males   |     |     |     |     |     |     |     | 8   |
9      | males   |     |     |     |     |     |     |     |     | 9
10     | females | 1   |     |     |     |     |     |     |     |
12     | females |     | 2   |     |     |     |     |     |     |
13     | females |     |     | 3   |     |     |     |     |     |
14     | females |     |     |     | 4   |     |     |     |     |
15     | females |     |     |     |     | 5   |     |     |     |
16     | females |     |     |     |     |     | 6   |     |     |
17     | females |     |     |     |     |     |     | 7   |     |
18     | females |     |     |     |     |     |     |     | 8   |
20     | females |     |     |     |     |     |     |     |     | 9

</small>

## Data {.smaller}

```{r, echo=FALSE}
data <- read.csv(file = "ANOVA_mixed.csv", header = T)

data[,-1] <- round(data[,-1],2)

datatable(data, rownames = FALSE, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```


