---
title: "Statistical Reasoning"
subtitle: "Null Hypothesis Testisn"
author: "Klinkenberg"
date: "06 sep 2023"
footer: "SMCR / SMCO"
format: 
  html: default
  revealjs: 
    output-ext: slide.html
---

## Course structure

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# install.packages('googleVis')
# install.packages('timeline')
# install.packages("readxl")
# library("timeline")
library("googleVis")
library("readxl")
```

```{r, echo=FALSE, eval=TRUE, results='asis', warning=FALSE, message=FALSE}

bachelor.psy <- read_xlsx("course_structure.xlsx", sheet = 5)

Timeline <- gvisTimeline(bachelor.psy, barlabel = "naam", rowlabel = "onderdeel",
                         start = "start", 
                         end = "end",
                         options  = list(timeline="{groupByRowLabel : true,
                                                    colorByRowLabel : true}",
                                         width    = 1000, 
                                         height   = 900)
                         )

cat(Timeline$html$chart)
```

## Course parts

* Lectures: On campus / Online / Video recording
* Preparatory Assignment: Submit in Testvision and Reflection in canvas
* Tutorials: For your support and inclass assessment
* Exam: Knowledge and skills

## Grading

$$\text{Final grade} = 0.9 \times \text{exam grade} + 0.1 \times \text{credit points}$$

* Exam
* Credits for
  * Tutorial participation
  * Preparatory assignment

## Credits {.smaller}

> You get 10 at the start of the course.

Number of times missed | Deduction for TU |	Deduction for PA
-----------------------|------------------|-----------------
1x                     | 0                | 0
2x                     | 0	              | 2
3x                     | 2                | 3
4x                     | 4                | 4
5x or more             | 5                | 5

See canvas for all the details.

## Learning

<iframe style="height:400px;  width:100%;" src="https://www.youtube.com/embed/JC82Il2cjqA" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

<!-- Add the required topics below -->

```{r child="../../../../topics/statistical_reasoning/statistical_reasoning.qmd"}
```

```{r child="../../../../topics/empirical_cycle/empirical_cycle.qmd"}
```

```{r child="../../../../topics/coin_expreiment/coin_experiment.qmd"}
```

<!-- Footer insert below -->

```{r child="../../../../footer.qmd"}
```

