# Test Utility

<!-- Allow color highlighting in mathjax -->
<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { extensions: ["color.js"] }});
</script>

## What is Test Utility {.subsection}

Test Utility, litterally refers to using a test as a utility, as a tool.

### Costs and benefits

* monotary
* non-monotary

The relationship between the costs and benefits are determined by the psychometric properties of the test.

These properties can be emperically tested. We will show you the reasoning process in this lecture.

## Predictive / Criterion Validity {.subsection}

Often we want to predict based on testscores how an individual will perform at a later time.

* The thing we want to predict is called the **criterion**
    * Example: Selection of good employees or good students
* The test we use to predict is obviously called the **predictor**

The predictive validity is defined as the correlation between the **predictor** / testscore and the **criterion**.

**NOTE:** This is no different than the $R$ in regular regression analysis, the correlation between the outcome and the model.

## {data-background=../../../topics/test_utility/tile.jpg data-background-size=contain .flexbox .vcenter}

<p style="color:darkblue; font-size:200%; text-align: center;">A corelation of zero,<br>is no predicting hero</p> <!-- [Dutch Tile](https://www.flickr.com/photos/paulussie/15582506882) -->


## Utility (interactive)

```{r, child="utility_interactive_scatterplot.Rmd"}
```

## CASE STUDY:

Admission test (NL: selectie aan de poort)

* Testweek data from UvA Psychology students
    * Motivation
    * Study results
* The data was gathered before we actually had admission tests (all students where admitted).
* Because of this, we can determine the effectiveness of a supposed admissions procedure.
* We can calculate how much beter our results would be if we had actually selected.
* This comparison will show us how much ther is to gain and the price we have to pay

## Hit Rate {.subsection}

Percentage of correct conclusions.

$$\definecolor{green}{RGB}{0,128,0}
\text{Hit Rate} = \frac{{\color{green}\text{TP}} + \color{green}\text{TN}}{N}$$

## Base Rate {.subsection}

Percentage of correct conclusions when not selecting.

$$\definecolor{red}{RGB}{255,0,0}
\text{Base Rate} = \frac{{\color{green}\text{TP}} + \color{red}\text{FN}}{N}$$

## Hit Rate - Base Rate

To determine the benefit of selection we need to know what the hit rate is compared to the base rate.

$$\text{Hit Rate} - \text{Base Rate} = \text{Benefit}$$

## Sensitivity {.subsection}

We can determine the efficiency of selection by calculating the sensitivity and specificity of a test.

**Sensitivity:** Percentage of eligible candidates that will be selected.

$$\text{Sensitivity} = \frac{\color{green}\text{TP}}{{\color{red}\text{FN}} + \color{green}\text{TP}}$$

## Specificity {.subsection}

**Specificity:** Percentage of inapt candidates that will be rejected.

$$\text{Specificity} = \frac{\color{green}\text{TN}}{{\color{green}\text{TN}} + \color{red}\text{FP}}$$

## Taylor Russell Tables {.subsection}


[Taylor, H, and Russell, J. “The Relationship of Validity Coefficients to the Practical Effectiveness of Tests in Selection: Discussion and Tables.” Journal of Applied Psychology 23 (1939): n. pag. Web.](https://lib.uva.nl/permalink/31UKB_UAM1_INST/h37vvk/proquest1290354749)


##

<div style="font-size:30%;">

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# install.packages("readxl")
library("readxl")
library("DT")
tr.table <- read_excel("TaylorRusselTable.xlsx")
datatable(tr.table, extensions = 'FixedColumns', rownames = F,
  options = list(
    pageLength = 21,
    dom = 't',
    scrollX = TRUE,
    fixedColumns = TRUE
  ) )
```

</div>
