# ANCOVA

```{r, echo=FALSE}
# Clear memory
rm(list=ls())

if(!"ggplot2" %in% installed.packages()) { install.packages("ggplot2") }
library("ggplot2")

set.seed(1976)

if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## ANCOVA {.section}

Analysis of covariance (ANCOVA) is a general linear model which blends ANOVA and regression. ANCOVA evaluates whether population means of a dependent variable (DV) are equal across levels of a categorical independent variable (IV) often called a treatment, while statistically controlling for the effects of other continuous variables that are not of primary interest, known as covariates (CV).

[WIKIPEDIA](https://en.wikipedia.org/wiki/Analysis_of_covariance)

## Data example

We want to test the difference in national extraversion but want to controll for openness to experience.

* Dependent variable: Extraversion
* Independent variabele: Nationality
     + Dutch
     + German
     + Belgian
* Covariate: Openness to experience

## Simulate data {.smaller}

```{r}
# Simulate data
n = 20
k = 3
group = round(runif(n,1,k),0)
mu.covar    = 8
sigma.covar = 1
covar = round(rnorm(n,mu.covar,sigma.covar),2)

# Create dummy variables
dummy.1 <- ifelse(group == 2, 1, 0)
dummy.2 <- ifelse(group == 3, 1, 0)

# Set parameters
b.0 = 15 # initial value for group 1
b.1 = 3  # difference between group 1 and 2
b.2 = 4  # difference between group 1 and 3
b.3 = 3  # Weight for covariate

# Create error
error = rnorm(n,0,1)
```

## Define the model {.subsection}

$${outcome} = {model} + {error}$$
$${model} = {indvar} + {covariate} = {nationality} + {openness}$$

Formal model

$$b_0 + b_1 {dummy}_1 + b_2 {dummy}_2 + b_3 covar$$


```{r}
# Define model
outcome = b.0 + b.1 * dummy.1 + b.2 * dummy.2 + b.3 * covar + error
```

## Dummies {.smaller}

```{r, echo=FALSE}
outcome <- round(outcome,2)
error <- round(error,2)

n <- 1:length(outcome)

dummies <- data.frame(group, b.0, b.1, dummy.1, b.2, dummy.2, b.3, covar, error, outcome)

datatable(dummies, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## The data

```{r, echo=FALSE}
# put data in data frame
data <- data.frame(group, covar, outcome)

# Order by group
data <- data[order(data$group),]
data <- cbind(n, data)
# Write data for use in SPSS
write.csv(data, "ANCOVA.csv", row.names=FALSE)

datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Group means {.subsection}

```{r}
aggregate(outcome ~ group, data, mean)
```

## Model fit no covar {.smaller .subsection}

What are the beta coëfficients based on the data without the covariate?

```{r}
fit.group <- lm(outcome ~ factor(group), data); fit.group
fit.group$coefficients[2:3] + fit.group$coefficients[1]
```

$${Dutch} = `r  fit.group$coefficients[1]` \> {German} = `r fit.group$coefficients[2] + fit.group$coefficients[1]` \> {Belgian} = `r fit.group$coefficients[3] + fit.group$coefficients[1]`$$

## Model fit only covar {.subsection}

What is the weight of only the covariate?

```{r}
fit.covar <- lm(outcome ~ covar, data)
fit.covar
```

## Model fit with covar {.smaller .subsection}

```{r}
fit <- lm(outcome ~ factor(group) + covar, data); fit
fit$coefficients[2:3] + fit$coefficients[1]
```

```{r, echo=FALSE}
fit$coefficients <- round(fit$coefficients, 2)
```

$${Dutch} = `r  fit$coefficients[1]` \> {German} = `r fit$coefficients[2] + fit$coefficients[1]` \> {Belgian} = `r fit$coefficients[3] + fit$coefficients[1]`$$

## Total variance {.subsection}

What is the total variance

$${MS}_{total} = s^2_{outcome} = \frac{{SS}_{outcome}}{{df}_{outcome}}$$

```{r}
ms.t = var(data$outcome); ms.t
ss.t = var(data$outcome) * (length(data$outcome) - 1); ss.t
```

## The data

```{r, echo=FALSE}
# Add grand mean to data frame
data$grand.mean <- mean(data$outcome)

datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Total variance visual {.subsection}

```{r}
plot <- ggplot(data, aes(x=n, y=outcome)) + geom_point(shape=1) +
        geom_hline(yintercept=mean(outcome)) +
        geom_segment(aes(x = n, y = grand.mean, xend = n, yend = outcome)) +
        ggtitle("Total variance")
plot
```

## Model variance group {.smaller}

The model variance consists of two parts. One for the independent variable and one for the covariate. Lets first look at the independent variable.

```{r, echo=FALSE}
data$model.group <- round(fit.group$fitted.values, 2)

datatable(data, options = list(searching = FALSE, scrollY = 315, paging = F, info = F))
```

## Model variance group visual {.subsection}

```{r}
plot + geom_segment(aes(x = n, y = grand.mean, xend = n, yend = data$model.group, colour = factor(group))) +
       ggtitle("Only group variance")
```

## Model variance covariate visual {.subsection}

```{r}
data$model.covar <- round(fit.covar$fitted.values,2)
plot + geom_segment(aes(x = n, y = grand.mean, xend = n, yend = data$model.covar, colour = factor(group))) +
       ggtitle("Only covariate variance")
```

## Model variance group and covariate {.smaller}

```{r, echo=FALSE}
data$model       <- round(fit$fitted.values, 2)

datatable(data, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Model variance group and covariate visual {.subsection}

```{r}
plot + geom_segment(aes(x = n, y = grand.mean, xend = n, yend = data$model, colour = factor(group))) + 
       ggtitle("Both group and covariate variance")
```

## Error variance with covariate {.subsection}

```{r}
plot + geom_segment(aes(x = n-.1, y = outcome, xend = n-.1, yend = data$model, colour = "purple")) + 
       ggtitle("Error variance")
```

## Sums of squares {.subsection}

```{r}
SS.model = with(data, sum((model       - grand.mean)^2))

# Sums of squares for individual effects
SS.model.group = with(data, sum((model.group - grand.mean)^2))
SS.model.covar = with(data, sum((model.covar - grand.mean)^2))

SS.covar = SS.model - SS.model.group; SS.covar ## SS.covar corrected for group
SS.group = SS.model - SS.model.covar; SS.group ## SS.group corrected for covar
```

## F-ratio {.subsection}

$$F = \frac{{MS}_{model}}{{MS}_{error}} = \frac{{SIGNAL}}{{NOISE}}$$

## Adjusted means {.smaller .subsection}

```{r}
# Add dummy variables
data$dummy.1 <- ifelse(data$group == 2, 1, 0)
data$dummy.2 <- ifelse(data$group == 3, 1, 0)

# b coefficients
b.cov = fit$coefficients["covar"];          b.int = fit$coefficients["(Intercept)"]
b.2   = fit$coefficients["factor(group)2"]; b.3   = fit$coefficients["factor(group)3"]

# Adjustment factor for the means of the independent variable
data$mean.adj <- with(data, b.int + b.cov * mean(covar) + b.2 * dummy.1 + b.3 * dummy.2)

aggregate(mean.adj ~ group, data, mean)
```

## Real $\beta$'s

```{r}
b.0 = 15 # initial value for group 1
b.1 = 3  # difference between group 1 and 2
b.2 = 4  # difference between group 1 and 3
b.3 = 3  # Weight for covariate

cbind(m.covar = mu.covar*3, 
      BETA    = c(b.0, b.0+b.1, b.0+b.2), 
      sum     = mu.covar*3 + c(b.0, b.0+b.1, b.0+b.2))
```



