# ANOVA<br>One-way repeated

Compare subjects on 2 or more repeated measures

```{r, echo=FALSE}
rm(list=ls())
if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Example

Measure driving ability in a driving simulator. Test in three conditions.

* Alcohol none
* Alcohol some
* Alcohol much

## The data {.smaller}

```{r, echo=FALSE}
data <- read.csv("ANOVA REP alcohol_speed.csv")
data <- data[,1:4]
data <- round(data[, 1:4], 3)
names(data) <- c('pp', 'none_alc', 'some_alc', 'much_alc')
datatable(data, rownames = FALSE, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Formulas {.smaller}

Variance | Sum of Squares | df | Mean Squares | F-ratio
---------|----------------|----|--------------|---------
Between  | ${SS}_{{between}} = {SS}_{{total}} - {SS}_{{within}}$ | ${DF}_{{total}}-{DF}_{{within}}$ | $\frac{{SS}_{{between}}}{{DF}_{{between}}}$ | &nbsp;
Within  | ${SS}_{{within}} = \sum{s_i^2(n_i-1)}$ | $(n_i-1)n$ | $\frac{{SS}_{{within}}}{{DF}_{{within}}}$ | &nbsp;
• Model | ${SS}_{{model}} = \sum{n_k(\bar{X}_k-\bar{X})^2}$ | $k-1$ | $\frac{{SS}_{{model}}}{{DF}_{{model}}}$ | $\frac{{MS}_{{model}}}{{MS}_{{error}}}$ | &nbsp;
• Error  | ${SS}_{{error}} = {SS}_{{within}} - {SS}_{{model}}$ | $(n-1)(k-1)$ | $\frac{{SS}_{{error}}}{{DF}_{{error}}}$ | &nbsp;
Total | ${SS}_{{total}} = s_{grand}^2(N-1)$ | $N-1$ | $\frac{{SS}_{{total}}}{{DF}_{{total}}}$ | &nbsp;

Where $n_i$ is the number of observations per person and $k$ is the number of conditions. These two are equal for a one-way repeated ANOVA. Furthermore $n$ is the number of subjects per condition and $N$ is the total number of data points $n \times k$.

## ${MS}_{total}$

```{r}
# Assign to individual variables
none_alc = data$none_alc
some_alc = data$some_alc
much_alc = data$much_alc
total    = c(none_alc,some_alc,much_alc)
```

$${MS}_{total} = \frac{{SS}_{{total}}}{{DF}_{{total}}} = s_{grand}^2$$

```{r}
MS_total = var(total); MS_total
```

## ${SS}_{total}$

$${DF_{total}} = N-1 \\
{SS}_{{total}} = s_{grand}^2(N-1)$$

```{r}
N = length(total)
DF_total = N - 1
SS_total = MS_total * DF_total; SS_total

sum((total - mean(total))^2)
```

## ${SS}_{total}$ visual {.smaller}

```{r}
# Set offset
offset = .2

# Calculate n
n = length(none_alc)

# Create plot
plot(none_alc,
     xlab = 'participants',
     ylab = 'Brokken',
     xlim = c(.5,22),
     ylim = c(3,7.5),
     col='green')
points((1:n)-offset, some_alc, col='red')
points((1:n)+offset, much_alc,   col='blue')

# Add the total mean
lines(c((1-offset),(n+offset)),rep(mean(total),2),col='black',lwd=2)

segments(1:n,        mean(total), 1:n,        none_alc)
segments(1:n-offset, mean(total), 1:n-offset, some_alc)
segments(1:n+offset, mean(total), 1:n+offset, much_alc)

text(n+offset,mean(total),expression(bar(X)[grand]),pos=4)

p <- recordPlot()
```

-------

```{r}
p

# add grid
segments(1:21-.5, 3, 1:21-.5, 7.5, col='grey')


p <- recordPlot()
```

## ${MS}_{within}$

$${MS}_{within} = \frac{{SS}_{{within}}}{{DF}_{{within}}} \\
{DF}_{within} = (n_i-1)n$$

```{r}
n.i = 3  # Number of mesurements per individual (none, some, much)
n   = 20 # Number of mesurements per group

DF_within = (n.i - 1) * n
DF_within
```

## ${SS}_{within}$ {.smaller}

$${SS}_{{within}} = \sum{s_i^2(n_i-1)}$$

```{r}
var_pp = apply(cbind(none_alc, some_alc, much_alc),1,var)
ss_pp  = var_pp * (n.i - 1)

SS_within = sum(ss_pp); SS_within

mean_pp = apply(cbind(none_alc, some_alc, much_alc),1,mean)

sum(c((none_alc   - mean_pp)^2, 
      (some_alc - mean_pp)^2,
      (much_alc   - mean_pp)^2))
```

## ${SS}_{within}$ data

```{r, echo=FALSE}
data$mean_pp <- round(mean_pp,3)
data$ss_pp <- round(ss_pp, 3)
datatable(data, rownames = FALSE, options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## ${SS}_{within}$ visual

```{r}
p
# Personal means
segments(1:n-offset, mean_pp, 1:n+offset, mean_pp, col='orange', lwd=3)
# The lines show the within deviation from the personal mean.
segments(1:n-offset, mean_pp, 1:n-offset, some_alc, col='orange', lwd=3)
segments(1:n, mean_pp,        1:n,        none_alc,   col='orange', lwd=3)
segments(1:n+offset, mean_pp, 1:n+offset, much_alc,   col='orange', lwd=3)

# Add personal means symbols
for(i in 1:n) { text(i-(offset/2), mean_pp[i], expression(bar(X)[i]), pos=2, cex=.5)}

p <- recordPlot()
```

## ${MS}_{between}$

$${MS}_{between} = \frac{{SS}_{{between}}}{{DF}_{{between}}}$$

$${DF}_{between}-{DF}_{{within}} \\
{SS}_{between} = {SS}_{total} - {SS}_{within}$$ 

```{r, eval=TRUE}
SS_between = SS_total - SS_within
SS_between

DF_between = DF_total - DF_within
DF_between
```

## ${MS}_{model}$

$${MS}_{model} = \frac{{SS}_{{model}}}{{DF}_{{model}}} \\
{DF}_{model} = k-1$$

```{r}
k = 3
DF_model = k - 1
DF_model
```

## ${SS}_{model}$

$${SS}_{model} = \sum{n_k(\bar{X}_k-\bar{X})^2}$$

```{r}
# SS model
n_k1 = length(none_alc)
n_k2 = length(some_alc)
n_k3 = length(much_alc)

# Calculate sums of squares for the model
SS_k1 = n_k1 * (mean(none_alc)   - mean(total))^2
SS_k2 = n_k2 * (mean(some_alc) - mean(total))^2
SS_k3 = n_k3 * (mean(much_alc)   - mean(total))^2

SS_model = sum(SS_k1, SS_k2, SS_k3)
SS_model
```

## ${SS}_{model}$ visual

```{r, eval=TRUE}
p
# Add the no alcohol mean
lines(c((1),(n)),rep(mean(none_alc),2),col='green',lwd=2)
text(n+offset,mean(none_alc),expression(bar(X)[none]),pos=4)

# With the bit alcohol mean
lines(c((1-offset),(n-offset)),rep(mean(some_alc),2),col='red',lwd=2)
text(n+offset,mean(some_alc),expression(bar(X)[some]),pos=4)

# With the much alcohol mean
lines(c((1+offset),(n+offset)),rep(mean(much_alc),2),col='blue',lwd=2)
text(n+offset,mean(much_alc),expression(bar(X)[much]),pos=4)

# The lines show the model deviation from the total mean.
segments(1:n,        mean(total), 1:n,        mean(none_alc),   col='green')
segments(1:n-offset, mean(total), 1:n-offset, mean(some_alc), col='red')
segments(1:n+offset, mean(total), 1:n+offset, mean(much_alc),   col='blue')

p <- recordPlot()
```

## ${MS}_{error}$

$$\frac{{SS}_{error}}{{DF}_{error}} \\
{DF}_{error} = (n-1)(k-1)$$

```{r}
DF_error = DF_within - DF_model
DF_error
```


## ${SS}_{error}$

$${SS}_{error} = {SS}_{within} - {SS}_{model}$$

```{r}
SS_error = SS_within - SS_model
SS_error
```

## F ratio

$$F = \frac{{MS}_{{model}}}{{MS}_{{error}}}$$

```{r}
# Calculate mean squares
MS_model = SS_model / DF_model
MS_error = SS_error / DF_error

# Calculate F statistic
F = MS_model / MS_error
F
```

## Visualize

```{r}
library('visualize')
visualize.f(F, DF_model, DF_error, section="upper")
```